---
title: "Gibbs Prior Joint"
format: html
editor: visual
---

# Setup

```{r}
setwd("/Users/brandonmarks/Desktop/Research/Palacios Lab/Untitled/Enertree/Brandon/gibbs_prior")
library("ape")
library("phylodyn")
library(phangorn)
library(phyclust)
library(phylotools)
library(fmatrix)
library(phangorn)
library(coda)
source("utils.R")
```

## Core Functions

```{r}

log_posterior_joint <- function(beta, M, coal_times, prior_sd, lambda = 1, start_tree = NULL, mode = "average") {
  tree_fmat <- sampleF(M, beta, iter = 2, diam = 1, startF = start_tree, take_every = 1)$chainF[[2]] ##SHOULD THIS RUN WHEN WE SAMOPLE M

  log_botlzman_dist <- Logdistance_prob(tree_fmat, M, beta = beta, diam = 1)
  if (mode == "average") {
    log_likelihood <- 0
  } else if (mode == "max") {
    log_likelihood <- -Inf
  } else {
    stop("Invalid mode. Choose 'average' or 'max'.")
  }

  R <- 10
  for (i in 1:R) {
    rooted_tree <- mytree_from_F(tree_fmat, coal_times)
    ll <- pml(rooted_tree, sequences)$log
    if (mode == "average") {
      log_likelihood <- log_likelihood + ll
    } else {
      log_likelihood <- max(log_likelihood, ll)
    }
  }
  if (mode == "average") {
    log_likelihood <- log_likelihood / R
  }

  n_tips <- ncol(M_true_fmat) + 1
  log_rooted_given_unrooted <- log(2^(num_cherries(tree_fmat)) / factorial(n_tips))
  log_prior <- dnorm(log(beta), mean = 0, sd = prior_sd, log = TRUE)
  log_posterior <- log_likelihood / lambda + log_rooted_given_unrooted + log_botlzman_dist + log_prior

  list(log_posterior = log_posterior, tree_fmat = tree_fmat)
}
```

```{r}
run_mh_beta_M <- function(
  M_init,
  beta_init,
  prior_sd,
  n_iter,
  coal_times,
  proposal_sd,
  lambda = 1,
  mode = "average",
  include_tree = FALSE,
  take_every = 1000
) {
  beta_chain <- numeric(n_iter)
  log_post_chain <- numeric(n_iter)
  M_chain <- vector("list", ceiling(n_iter / take_every))
  tree_chain <- vector("list", ceiling(n_iter / take_every))

  accepted_beta <- 0
  accepted_M <- 0

  current_beta <- beta_init
  current_g <- log(current_beta)
  current_M <- M_init

  current_res <- log_posterior_joint(
    current_beta, current_M, coal_times, prior_sd,
    lambda = lambda, start_tree = current_M, mode = mode
  )
  current_log_post <- current_res$log_posterior
  current_tree <- current_res$tree_fmat

  best_log_post <- current_log_post
  best_beta <- current_beta
  best_M <- current_M
  best_tree <- current_tree

  beta_chain[1] <- current_beta
  log_post_chain[1] <- current_log_post
  M_chain[[1]] <- current_M
  if (include_tree) {
    tree_chain[[1]] <- current_tree
  }

  for (iter in 2:n_iter) {
    proposed_g <- rnorm(1, mean = current_g, sd = proposal_sd)
    proposed_beta <- exp(proposed_g)

    beta_res <- log_posterior_joint(
      proposed_beta, current_M, coal_times, prior_sd,
      lambda = lambda, start_tree = current_tree, mode = mode
    )
    proposed_log_post <- beta_res$log_posterior

    if (log(runif(1)) < (proposed_log_post - current_log_post)) {
      current_beta <- proposed_beta
      current_g <- proposed_g
      current_log_post <- proposed_log_post
      current_tree <- beta_res$tree_fmat
      accepted_beta <- accepted_beta + 1
    }

    current_encod <- my_encod(current_M)
    proposed_M <- Fmat_from_myencod(proposal_myencod(current_encod))
    M_res <- log_posterior_joint(
      current_beta, proposed_M, coal_times, prior_sd,
      lambda = lambda, start_tree = current_tree, mode = mode
    )
    proposed_log_post_M <- M_res$log_posterior

    if (log(runif(1)) < (proposed_log_post_M - current_log_post)) {
      current_M <- proposed_M
      current_log_post <- proposed_log_post_M
      current_tree <- M_res$tree_fmat
      accepted_M <- accepted_M + 1
    }

    if (current_log_post > best_log_post) {
      best_log_post <- current_log_post
      best_beta <- current_beta
      best_M <- current_M
      best_tree <- current_tree
    }

    beta_chain[iter] <- current_beta
    log_post_chain[iter] <- current_log_post

    if (iter %% take_every == 0) {
      slot <- iter / take_every
      M_chain[[slot]] <- current_M
      if (include_tree) {
        tree_chain[[slot]] <- current_tree
      }
    }

    per10 <- max(1, floor(n_iter / 10))
    if (iter %% per10 == 0) {
      cat("Iteration:", iter, "Current Beta:", current_beta, "Log Posterior:", current_log_post, "\n")
    }
  }

  out <- list(
    beta_chain = beta_chain,
    log_posterior = log_post_chain,
    M_chain = M_chain,
    best_beta = best_beta,
    best_M = best_M,
    best_tree = best_tree,
    acceptance_rate_beta = accepted_beta / (n_iter - 1),
    acceptance_rate_M = accepted_M / (n_iter - 1)
  )

  valid_M_chain <- Filter(Negate(is.null), M_chain)
  if (length(valid_M_chain) > 0) {
    average_M_raw <- Reduce(`+`, valid_M_chain) / length(valid_M_chain)
    out$average_M_raw <- average_M_raw
    out$average_M <- nearby_Fmat(average_M_raw)
  } else {
    out$average_M_raw <- NULL
    out$average_M <- NULL
  }

  if (include_tree) {
    out$tree_chain <- tree_chain
  }

  out
}
```

## Data And Reference Objects

```{r}
n_tips <- 25
set.seed(123)
sequences <- as.phyDat(read.FASTA("sequences.fasta"))

M_true_tree <- read.tree("true_tree.newick")
inter_coal_times <- coalescent.intervals(M_true_tree)$interval.length
inter_coal_times[which(inter_coal_times <= 0.000001)] <- 0.0000001
true_coalescent_times <- cumsum(inter_coal_times)
M_true_fmat <- gen_Fmat(M_true_tree, tol = 8)
```

### UPGMA Reference

```{r}
dm <- dist.hamming(sequences)
upgma_tree <- upgma(dm)
upgma_tree <- update_time(upgma_tree, true_coalescent_times)
M_upgma_fmat <- gen_Fmat(upgma_tree, tol = 8)
```

### MLE Reference

```{r}
# Step 2: Compute a fully bifurcating starting tree (Neighbor-Joining)
dm <- dist.hamming(sequences)
start_tree <- nj(dm)
start_tree <- midpoint(start_tree)
start_tree <- multi2di(start_tree, random = FALSE)

# Step 3: Fit initial likelihood
fit <- pml(start_tree, data = sequences)

# Step 4: Optimize tree with MLE under strict clock
fit <- optim.pml(fit, model = "JC", rearrangement = "stochastic", optClock = TRUE)

# Step 5: Extract and repair the MLE tree
mle_tree <- fit$tree
mle_tree <- midpoint(mle_tree)
mle_tree <- multi2di(mle_tree, random = FALSE)
mle_tree <- update_time(mle_tree, true_coalescent_times)
min_branch <- 1e-8
mle_tree$edge.length[mle_tree$edge.length < min_branch] <- min_branch

M_mle_fmat <- gen_Fmat(mle_tree, tol = 8)
```

## Joint Chain (UPGMA Init)

```{r}
res_joint <- run_mh_beta_M(
  M_init = M_upgma_fmat,
  beta_init = 1,
  prior_sd = 1,
  n_iter = 10000,
  coal_times = true_coalescent_times,
  proposal_sd = 0.5,
  lambda = 1,
  include_tree = TRUE,
  mode = "max",
  take_every = 1
)

beta_chain_joint <- res_joint$beta_chain
hist(beta_chain_joint, breaks = 30, main = "Posterior of Beta (Joint Max Mode)", xlab = "Beta")
```

```{r}
best_M_tree <- res_joint$best_M
joint_avg_M <- res_joint$average_M
print(distance_Fmat(best_M_tree, M_true_fmat))
print(distance_Fmat(joint_avg_M, M_true_fmat))
```

## Joint Chain (True M Init)

```{r}
res_joint_true <- run_mh_beta_M(
  M_init = M_true_fmat,
  beta_init = 1,
  prior_sd = 1,
  n_iter = 10000,
  coal_times = true_coalescent_times,
  proposal_sd = 0.5,
  lambda = 1,
  include_tree = TRUE,
  mode = "max",
  take_every = 1
)

beta_chain_joint_true <- res_joint_true$beta_chain
hist(beta_chain_joint_true, breaks = 30, main = "Posterior of Beta (Joint Max Mode, True M Init)", xlab = "Beta")
```

## Shared BEAST Reference Samples

```{r}
library(progress)
trees_beast <- read.nexus("HKY1000Gibbs-sequences.trees")
N <- 10000
F.list <- vector("list", N)
pb <- progress_bar$new("[:bar] :percent", total = N)
for (i in 1:N) {
  name <- names(trees_beast)[i]
  pb$tick()
  F.list[[i]] <- gen_Fmat(trees_beast[[name]], tol = 8)
}
```

## Shared Subsampling

```{r}
burn_in <- 50
sub_samp <- 1000
chain_gen <- res_joint$tree_chain

burn_in <- min(burn_in, length(chain_gen) - 1)
chain_gen <- chain_gen[(burn_in + 1):length(chain_gen)]

set.seed(42)
chain_data <- F.list
idx <- sample(length(chain_data), sub_samp)
trees_data <- chain_data[idx]

idx <- sample(length(chain_gen), sub_samp)
trees_gen <- chain_gen[idx]
```

```{r}
burn_in_true <- 50
chain_gen_true <- res_joint_true$tree_chain

burn_in_true <- min(burn_in_true, length(chain_gen_true) - 1)
chain_gen_true <- chain_gen_true[(burn_in_true + 1):length(chain_gen_true)]

sub_samp_true <- min(sub_samp, length(chain_gen_true), length(trees_data))
set.seed(43)
idx <- sample(length(chain_gen_true), sub_samp_true)
trees_gen_true <- chain_gen_true[idx]
```

## UPGMA Init Diagnostics

```{r}
tree_MDS_comparison_plot(
  trees_gen,
  trees_data,
  M_true_fmat,
  M_tree = best_M_tree,
  title = "Joint MDS UPGMA Comparison (MAX MODE)",
  save_path = "Joint_MDS_UPGMA_comparison_max_mode.png"
)
```

```{r}
tree_MDS_comparison_plot(
  trees_gen,
  trees_data,
  M_true_fmat,
  M_tree = joint_avg_M,
  title = "Joint MDS UPGMA Comparison (MAX MODE, Avg M)",
  save_path = "Joint_MDS_UPGMA_comparison_avg_m.png"
)
```

```{r}
matrix_MDS_comparison_plot(
  matrices = list(M_true_fmat, res_joint$average_M, res_joint$best_M),
  labels = c("True M", "Average M", "Max M"),
  chain_matrices = res_joint$tree_chain,
  coal_times = true_coalescent_times,
  sequences = sequences,
  color_by_log_likelihood = TRUE,
  sub_samp = 1000,
  seed = 44,
  title = "UPGMA MDS Log Likelihood Comparison",
  save_path = "Joint_MDS_UPGMA_log_likelihood_comparison.png"
)
```

```{r}
matrix_MDS_comparison_plot(
  matrices = list(M_true_fmat, res_joint$average_M, res_joint$best_M),
  labels = c("True M", "Average M", "Max M"),
  chain_matrices = res_joint$M_chain,
  sub_samp = 1000,
  seed = 44,
  title = "UPGMA MDS Comparison: True vs Average vs Max M",
  save_path = "Joint_MDS_UPGMA_comparison_true_average_max.png"
)
```

### ESS Diagnostics

```{r}
matrix_chain_ess_heatmap(
  res_joint$M_chain,
  title = "UPGMA ESS Heatmap for M Chain",
  save_path = "Joint_UPGMA_M_chain_ESS_heatmap.png"
)
```

```{r}
matrix_chain_ess_heatmap(
  res_joint$tree_chain,
  normalize = TRUE,
  per_n_samples = 1000,
  title = "UPGMA ESS per 1000 Samples for Tree Chain",
  save_path = "Joint_UPGMA_tree_chain_ESS_per_1000_heatmap.png"
)
```

## BEAST ESS Reference

```{r}
library(progress)
trees_beast_all <- read.nexus("100THOUSANDTREES-sequences.trees")
N <- 100000
ALLTREES <- vector("list", N)
pb <- progress_bar$new("[:bar] :percent", total = N)
for (i in 1:N) {
  name <- names(trees_beast_all)[i]
  pb$tick()
  ALLTREES[[i]] <- gen_Fmat(trees_beast_all[[name]], tol = 8)
}
```

```{r}
matrix_chain_ess_heatmap(
 ALLTREES,
   title = "ESS Heatmap for BEAST Trees",
  save_path = "Joint_BEAST_ESS_heatmap.png"
)
```

```{r}
matrix_chain_ess_heatmap(
  ALLTREES,
  normalize = TRUE,
  per_n_samples = 1000,
  title = "ESS per 1000 Samples for BEAST",
  save_path = "Joint_BEAST_ESS_per_1000_heatmap.png"
)
```

## Histogram And True-Init Comparisons

```{r}
tree_histogram_comparison_plot(
  chain_gen,
  chain_data,
  M_true_fmat,
  title = "UPGMA Joint Histogram Comparison (MAX MODE)",
  save_path = "Joint_UPGMA_Histogram_comparison_max_mode.png"
)
```

```{r}
tree_MDS_comparison_plot(
  trees_gen_true,
  trees_data,
  M_true_fmat,
  M_tree = M_true_fmat,
  title = "Joint MDS Comparison (MAX MODE, True M Init)",
  save_path = "Joint_MDS_comparison_true_init.png"
)
```

```{r}
tree_histogram_comparison_plot(
  chain_gen_true,
  chain_data,
  M_true_fmat,
  title = "Joint Histogram Comparison (MAX MODE, True M Init)",
  save_path = "Joint_Histogram_comparison_true_init.png"
)
```

## Joint Chain (MLE Init)

```{r}
res_joint_mle <- run_mh_beta_M(
  M_init = M_mle_fmat,
  beta_init = 1,
  prior_sd = 1,
  n_iter = 10000,
  coal_times = true_coalescent_times,
  proposal_sd = 0.5,
  lambda = 1,
  include_tree = TRUE,
  mode = "max",
  take_every = 1
)

beta_chain_joint_mle <- res_joint_mle$beta_chain
hist(beta_chain_joint_mle, breaks = 30, main = "Posterior of Beta (Joint Max Mode, MLE Init)", xlab = "Beta")
```

```{r}
best_M_tree_mle <- res_joint_mle$best_M
joint_avg_M_mle <- res_joint_mle$average_M
print(distance_Fmat(best_M_tree_mle, M_true_fmat))
print(distance_Fmat(joint_avg_M_mle, M_true_fmat))
```

### MLE Init Subsampling

```{r}
burn_in_mle <- 50
chain_gen_mle <- res_joint_mle$tree_chain

burn_in_mle <- min(burn_in_mle, length(chain_gen_mle) - 1)
chain_gen_mle <- chain_gen_mle[(burn_in_mle + 1):length(chain_gen_mle)]

sub_samp_mle <- min(sub_samp, length(chain_gen_mle), length(trees_data))
set.seed(45)
idx <- sample(length(chain_gen_mle), sub_samp_mle)
trees_gen_mle <- chain_gen_mle[idx]
```

### MLE Init Diagnostics

```{r}
tree_MDS_comparison_plot(
  trees_gen_mle,
  trees_data,
  M_true_fmat,
  M_tree = best_M_tree_mle,
  title = "Joint MDS Comparison (MAX MODE, MLE Init)",
  save_path = "Joint_MLE_init_MDS_comparison_max_mode.png"
)
```

```{r}
tree_MDS_comparison_plot(
  trees_gen_mle,
  trees_data,
  M_true_fmat,
  M_tree = joint_avg_M_mle,
  title = "Joint MDS Comparison (MAX MODE, MLE Init, Avg M)",
  save_path = "Joint_MLE_init_MDS_comparison_avg_m.png"
)
```

```{r}
matrix_MDS_comparison_plot(
  matrices = list(M_true_fmat, res_joint_mle$average_M, res_joint_mle$best_M),
  labels = c("True M", "Average M", "Max M"),
  chain_matrices = res_joint_mle$M_chain,
  sub_samp = 1000,
  seed = 46,
  title = "MDS Comparison: True vs Average vs Max M (MLE Init)",
  save_path = "Joint_MLE_init_MDS_true_average_max.png"
)
```

### MLE Init ESS Diagnostics

```{r}
matrix_chain_ess_heatmap(
  res_joint_mle$M_chain,
  title = "ESS Heatmap for M Chain (MLE Init)",
  save_path = "Joint_MLE_init_M_chain_ESS_heatmap.png"
)
```

```{r}
matrix_chain_ess_heatmap(
  res_joint_mle$tree_chain,
  normalize = TRUE,
  per_n_samples = 1000,
  title = "ESS per 1000 Samples for Tree Chain (MLE Init)",
  save_path = "Joint_MLE_init_tree_chain_ESS_per_1000_heatmap.png"
)
```

### MLE Init Histogram Comparison

```{r}
tree_histogram_comparison_plot(
  chain_gen_mle,
  chain_data,
  M_true_fmat,
  title = "Joint Histogram Comparison (MAX MODE, MLE Init)",
  save_path = "Joint_MLE_init_Histogram_comparison_max_mode.png"
)
```

## Joint Chain (UPGMA Temperature)

```{r}
res_joint_temp <- run_mh_beta_M(
  M_init = M_upgma_fmat,
  beta_init = 1,
  prior_sd = 1,
  n_iter = 10000,
  coal_times = true_coalescent_times,
  proposal_sd = 0.5,
  lambda = 10,
  include_tree = TRUE,
  mode = "max",
  take_every = 1
)

beta_chain_joint_temp <- res_joint_temp$beta_chain
hist(beta_chain_joint_temp, breaks = 30, main = "Posterior of Beta (Joint Max Mode)", xlab = "Beta")
```

```{r}
best_M_tree_temp <- res_joint_temp$best_M
joint_avg_M_temp <- res_joint_temp$average_M
print(distance_Fmat(best_M_tree_temp, M_true_fmat))
print(distance_Fmat(joint_avg_M_temp, M_true_fmat))
```

## Shared Subsampling

```{r}
burn_in <- 50
sub_samp <- 1000
chain_gen_temp <- res_joint_temp$tree_chain

burn_in <- min(burn_in, length(chain_gen_temp) - 1)
chain_gen_temp <- chain_gen_temp[(burn_in + 1):length(chain_gen_temp)]

idx <- sample(length(chain_gen_temp), sub_samp)
trees_gen_temp <- chain_gen_temp[idx]
```

## UPGMA Init Diagnostics

```{r}
tree_MDS_comparison_plot(
  trees_gen_temp,
  trees_data,
  M_true_fmat,
  M_tree = best_M_tree_temp,
  title = "Joint MDS UPGMA Temperature Comparison (MAX MODE)",
  save_path = "Joint_MDS_UPGMA_Temperature_comparison_max_mode.png"
)
```

```{r}
tree_MDS_comparison_plot(
  trees_gen_temp,
  trees_data,
  M_true_fmat,
  M_tree = joint_avg_M_temp,
  title = "Joint MDS UPGMA Temperature Comparison (MAX MODE, Avg M)",
  save_path = "Joint_MDS_UPGMA_Temperature_comparison_avg_m.png"
)
```

```{r}
matrix_MDS_comparison_plot(
  matrices = list(M_true_fmat, res_joint$average_M, res_joint$best_M),
  labels = c("True M", "Average M", "Max M"),
  chain_matrices = res_joint_temp$M_chain,
  sub_samp = 1000,
  seed = 44,
  title = "UPGMA MDS Temperature Comparison: True vs Average vs Max M",
  save_path = "Joint_MDS_UPGMA_Temperature_comparison_true_average_max.png"
)
```

### ESS Diagnostics

```{r}
matrix_chain_ess_heatmap(
  res_joint_temp$M_chain,
  title = "UPGMA Temperature ESS Heatmap for M Chain",
  save_path = "Joint_UPGMA_Temperature_M_chain_ESS_heatmap.png"
)
```

```{r}
matrix_chain_ess_heatmap(
  res_joint_temp$tree_chain,
  normalize = TRUE,
  per_n_samples = 1000,
  title = "UPGMA Temperature ESS per 1000 Samples for Tree Chain",
  save_path = "Joint_UPGMA_Temperature_tree_chain_ESS_per_1000_heatmap.png"
)
```
