---
title: "Finding Beta is Hard — All Experiments"
format:
  html:
    toc: true
    df-print: paged
editor: visual
execute:
  warning: false
  message: false
---

## Setup

```{r}
library(fmatrix)
library(dplyr)
library(tidyr)
library(purrr)
library(tibble)
library(ggplot2)


Logdistance_prob <- function(fmat, centre=NULL, beta=1, diam=1){
  n <- dim(fmat)[1] + 1
  if (is.null(centre)) centre <- kingman_m(n)
  -(beta/diam) * distance_Fmat(fmat, centre, dist="l2")^2
}

sampleF <- function(M, beta, iter, diam, startF = NULL){
  acceptance_rate <- 0
  n <- nrow(M) + 1

  chainF <- vector("list", iter)

  if (is.null(startF)) {
    current_tree <- rcoal(n)
    chainF[[1]]  <- gen_Fmat(current_tree, tol=13)
  } else {
    chainF[[1]] <- startF
  }

  current_Encod <- my_encod(chainF[[1]])
  f_current     <- Fmat_from_myencod(current_Encod)
  Logdist_curr  <- Logdistance_prob(f_current, M, beta=beta, diam=diam)

  MeanM <- matrix(0, nrow=n-1, ncol=n-1)

  for (j in 1:iter){
    proposed    <- proposal_myencod(current_Encod)
    f_proposed  <- Fmat_from_myencod(proposed)
    Logdist_prop <- Logdistance_prob(f_proposed, M, beta=beta, diam=diam)

    prob <- exp(Logdist_prop - Logdist_curr)

    if (runif(1) < prob){
      chainF[[j]]     <- f_proposed
      current_Encod   <- proposed
      f_current       <- f_proposed
      Logdist_curr    <- Logdist_prop
      acceptance_rate <- acceptance_rate + 1
    } else {
      chainF[[j]] <- f_current
    }

    MeanM <- MeanM + chainF[[j]]
  }

  MeanM <- MeanM/iter
  list(
    chainF = chainF,
    acceptance_rate = acceptance_rate/iter,
    MeanM = MeanM
  )
}


find_beta <- function(beta, CD_params, n=25, iter=100000, start_at_tree=FALSE, use_true_mean = TRUE, return_samples = FALSE){
  
  print(paste0("Testing beta = ", beta))
  
  burn_in     <- CD_params$burn_in
  k           <- CD_params$k
  trials      <- CD_params$trials
  batch_size  <- CD_params$batch_size
  print_every <- CD_params$print_every
  step0       <- CD_params$step0

  beta_star   <- CD_params$beta_star
  diam <- 1

  diam <- 1
  sample_tree <- rcoal(n)
  M <- gen_Fmat(sample_tree)
  if(start_at_tree){
    samples <-sampleF(M,beta,iter,diam, startF = M)
  }else{
    samples <-sampleF(M,beta,iter,diam)
  }

  chain <-samples$chainF
  burn_in <- min(burn_in, length(chain) - 1)
  chain <- chain[(burn_in+1):length(chain)]
  n_data <- length(chain)
  
  if(use_true_mean){
    distro_mean <- M
  }else{
    arr <- simplify2array(chain)
    distro_mean <- apply(arr, c(1, 2), mean)
  }
    

  beta_star_list <- numeric(trials)
  
  for (t in 1:trials){
    idx <- sample.int(n_data, batch_size, replace = TRUE)

    s_pos_sum <- 0
    s_neg_sum <- 0

    for (b in 1:batch_size){
      x0 <- chain[[ idx[b] ]]

      d0 <- distance_Fmat(x0, distro_mean, dist = "l2")
      s_pos_sum <- s_pos_sum + d0^2

      out <- sampleF(M = distro_mean, beta = beta_star, iter = k + 1, diam = diam, startF = x0)
      xk <- out$chainF[[ k + 1 ]]

      dk <- distance_Fmat(xk, distro_mean, dist = "l2")
      s_neg_sum <- s_neg_sum + dk^2
    }

    E_pos <- s_pos_sum / batch_size
    E_neg <- s_neg_sum / batch_size
    g_beta <- (E_neg - E_pos) / diam

    step <- step0
    beta_star <- beta_star + step * g_beta
    if (beta_star < 1e-8) beta_star <- 1e-8

    beta_star_list[t] <- beta_star

    #if (t %% print_every == 0){
      #cat("trial", t, "beta_star", beta_star, "\n")
    #}
  }

  out_tbl <- tibble(
    beta = beta,
    trial = 1:trials,
    beta_star = beta_star_list
  )

  if (return_samples) {
    return(list(
      TrueM = M,
      out_tbl    = out_tbl,
      samples = samples,
      chain   = chain
    ))
  }

  out_tbl
  
}

summarize_run <- function(traj, tail_frac = 0.5){
  # summarize a single trajectory tibble from find_beta()
  T <- nrow(traj)
  start <- floor(T*(1-tail_frac)) + 1
  tibble(
    beta = traj$beta[1],
    trials = T,
    beta_star_final = traj$beta_star[T],
    beta_star_mean_tail = mean(traj$beta_star[start:T]),
    beta_star_sd_tail = sd(traj$beta_star[start:T])
  )
}



full_process <- function(iter, n, name = summary_table, true_mean = TRUE){
  
  
  exp_grid <- tibble::tribble(
  ~beta, ~beta_star0, ~step0,
   5.0,       6,      0.10,
   3.0,       4,      0.10,
   2.0,       1,      0.10,
   1.0,       1,      0.05,
   0.9,       1,      0.03,
   0.7,       1,      0.03,
   0.5,       1,      0.01,
   0.3,       1,      0.01,
   0.1,       1,      0.01
)

base_params <- list(
  burn_in = 500,
  k = 8,
  trials = 100000,
  batch_size = 10,
  print_every = 1000
)

# ---- Run everything ----
set.seed(1)

runs <- exp_grid %>%
  mutate(
    CD_params = pmap(list(beta_star0, step0), ~modifyList(base_params, list(beta_star = ..1, step0 = ..2))),
    t0 = Sys.time(),
    traj = map2(beta, CD_params, ~find_beta(beta = .x, CD_params = .y, n = n, iter = iter, start_at_tree = FALSE, use_true_mean = true_mean)),
    t1 = Sys.time(),
    seconds = as.numeric(difftime(t1, t0, units = "secs")),
    summary = map(traj, summarize_run)
  )

# Summary table
summary_tbl <- runs %>%
  select(beta_star0, step0, seconds, summary) %>%
  unnest(summary) %>%
  relocate(beta, .before = everything()) %>%
  arrange(desc(beta))

summary_tbl
write.csv(summary_tbl, "summary_table_trueM_10000_n10.csv", row.names = FALSE)


print(knitr::kable(summary_tbl, digits = 4, caption = "Beta experiments (one run per beta). Tail stats use the last 50%"))
  
  
  
}
```

# N = 10

## Run all experiments (TRUE MEAN)

### iter = 10000

```{r}
iter = 10000
n = 10
full_process(iter = iter, n = n, name = "summary_table_trueM_iter_10000_n10.csv", true_mean = TRUE)


```

### iter = 50000

```{r}
iter = 50000
n = 10
full_process(iter = iter, n = n, name = "summary_table_trueM_iter_50000_n10.csv", true_mean = TRUE)


```

### iter = 100000

```{r}
iter = 100000
n = 10
full_process(iter = iter, n = n, name = "summary_table_trueM_iter_100000_n10.csv", true_mean = TRUE)


```

## Run all experiments (Infered MEAN)

### iter = 10000

```{r}
iter = 10000
n = 10
full_process(iter = iter, n = n, name = "summary_table_inferM_iter_10000_n10.csv", true_mean = FALSE)


```

### iter = 50000

```{r}
iter = 50000
n = 10
full_process(iter = iter, n = n, name = "summary_table_inferM_iter_50000_n10.csv", true_mean = FALSE)


```

### iter = 100000

```{r}
iter = 100000
n = 10
full_process(iter = iter, n = n, name = "summary_table_inferM_iter_100000_n10.csv", true_mean = FALSE)


```




# N = 20

## Run all experiments (TRUE MEAN)

### iter = 10000

```{r}
iter = 10000
n = 20
full_process(iter = iter, n = n, name = "summary_table_trueM_iter_10000_n20.csv", true_mean = TRUE)


```

### iter = 50000

```{r}
iter = 50000
n = 20
full_process(iter = iter, n = n, name = "summary_table_trueM_iter_50000_n20.csv", true_mean = TRUE)


```

### iter = 100000

```{r}
iter = 100000
n = 20
full_process(iter = iter, n = n, name = "summary_table_trueM_iter_100000_n20.csv", true_mean = TRUE)


```

## Run all experiments (Infered MEAN)

### iter = 10000

```{r}
iter = 10000
n = 20
full_process(iter = iter, n = n, name = "summary_table_inferM_iter_10000_n20.csv", true_mean = FALSE)


```

### iter = 50000

```{r}
iter = 50000
n = 20
full_process(iter = iter, n = n, name = "summary_table_inferM_iter_50000_n20.csv", true_mean = FALSE)


```

### iter = 100000

```{r}
iter = 100000
n = 20
full_process(iter = iter, n = n, name = "summary_table_inferM_iter_100000_n20.csv", true_mean = FALSE)


```




# N = 25

## Run all experiments (TRUE MEAN)

### iter = 10000

```{r}
iter = 10000
n = 25
full_process(iter = iter, n = n, name = "summary_table_trueM_iter_10000_n25.csv", true_mean = TRUE)


```

### iter = 50000

```{r}
iter = 50000
n = 25
full_process(iter = iter, n = n, name = "summary_table_trueM_iter_50000_n25.csv", true_mean = TRUE)


```

### iter = 100000

```{r}
iter = 100000
n = 25
full_process(iter = iter, n = n, name = "summary_table_trueM_iter_100000_n25.csv", true_mean = TRUE)


```

## Run all experiments (Infered MEAN)

### iter = 10000

```{r}
iter = 10000
n = 25
full_process(iter = iter, n = n, name = "summary_table_inferM_iter_10000_n25.csv", true_mean = FALSE)


```

### iter = 50000

```{r}
iter = 50000
n = 25
full_process(iter = iter, n = n, name = "summary_table_inferM_iter_50000_n25.csv", true_mean = FALSE)


```

### iter = 100000

```{r}
iter = 100000
n = 25
full_process(iter = iter, n = n, name = "summary_table_inferM_iter_100000_n25.csv", true_mean = FALSE)


```


# N = 40

## Run all experiments (TRUE MEAN)

### iter = 10000

```{r}
iter = 10000
n = 40
full_process(iter = iter, n = n, name = "summary_table_trueM_iter_10000_n40.csv", true_mean = TRUE)


```

### iter = 50000

```{r}
iter = 50000
n = 40
full_process(iter = iter, n = n, name = "summary_table_trueM_iter_50000_n40.csv", true_mean = TRUE)


```

### iter = 100000

```{r}
iter = 100000
n = 40
full_process(iter = iter, n = n, name = "summary_table_trueM_iter_100000_n40.csv", true_mean = TRUE)


```

## Run all experiments (Infered MEAN)

### iter = 10000

```{r}
iter = 10000
n = 40
full_process(iter = iter, n = n, name = "summary_table_inferM_iter_10000_n40.csv", true_mean = FALSE)


```

### iter = 50000

```{r}
iter = 50000
n = 40
full_process(iter = iter, n = n, name = "summary_table_inferM_iter_50000_n40.csv", true_mean = FALSE)


```

### iter = 100000

```{r}
iter = 100000
n = 40
full_process(iter = iter, n = n, name = "summary_table_inferM_iter_100000_n40.csv", true_mean = FALSE)


```